---
title: Test BiocNeighbors for UCell score smoothing
author: 
- Massimo Andreatta^[massimo.andreatta@unil.ch]
- Santiago Carmona^[santiago.carmona@unil.ch]
date: "03/08/2022"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'SmoothNN.html'))})
#output: html_notebook
---


```{r message=F, warning=F}
#renv::restore()

#BiocManager::install("UCell")
#remotes::install_github("carmonalab/UCell", ref="dev")  #development version

library(UCell)
```

#Object at: https://drive.switch.ch/index.php/s/3kM5PQ0tQaG6d6A
```{r}
library(ggplot2)
library(Seurat)
library(SeuratDisk)

reduction = "wnn.umap"
#reduction = "umap"

pbmc <- readRDS("data/pbmc_multimodal.downsampled20k.Tcell.seurat.RNA.rds")
#pbmc <- LoadH5Seurat("../Azimuth/data/pbmc_multimodal.h5seurat")

DimPlot(object = pbmc, reduction = "wnn.umap", group.by = "celltype.l2", label = TRUE, label.size = 3, repel = TRUE)
DimPlot(object = pbmc, reduction = "umap", group.by = "celltype.l2", label = TRUE, label.size = 3, repel = TRUE)
```

#Get some gene sets
```{r fig.height=6}
signatures <- list()
signatures$CD4T <- c("CD4","CD40LG")
signatures$CD8T <- c("CD8A","CD8B")
signatures$Treg <- c("FOXP3","IL2RA")
signatures$MAIT <- c("TRAV1-2","SLC4A10")
signatures$Tcell_gd <- c("TRDC", "TRGC1", "TRGC2", "TRDV1","TRAC-","TRBC1-","TRBC2-")
signatures$NK <- c("KLRD1","NKG7","NCR1","FCGR3A","CD3D-","CD3E-","CD3G-","CD8A-","CD8B-")

t1 <- Sys.time()
pbmc <- AddModuleScore_UCell(pbmc, features = signatures, name = NULL, ncores=8)
t2 <- Sys.time()
t2 - t1

FeaturePlot(pbmc, reduction = reduction,
            features = c("CD4T","CD8T","Tcell_gd","MAIT"), max.cutoff = 'q99') &
  theme(aspect.ratio = 1)

```

Find NN
```{r}
library(BiocNeighbors)
k <- 20

space <- pbmc@reductions$pca@cell.embeddings
#space <- pbmc@reductions$wnn.umap@cell.embeddings

nn <- findKNN(space, k=k, BNPARAM=AnnoyParam())
```

Time benchmark
```{r eval=F}
library(BiocNeighbors)
k <- 50

space <- pbmc@reductions$pca@cell.embeddings

t1 <- Sys.time()
#Method1
nn <- findKNN(space, k=k, BNPARAM=AnnoyParam())
t2 <- Sys.time()
t2 - t1


#Method2
t1 <- Sys.time()
nn <- findKNN(space, k=k, BNPARAM=HnswParam())
t2 <- Sys.time()
t2 - t1

#Multi-core?
t1 <- Sys.time()
BPPARAM <- BiocParallel::MulticoreParam(workers = 8)
#Method1
nn <- findKNN(space, k=k, BNPARAM=AnnoyParam(), BPPARAM=BPPARAM)
t2 <- Sys.time()
t2 - t1
```

Did it work?
```{r}
DimPlot(pbmc, reduction = reduction, cells.highlight = nn$index[1,], sizes.highlight = 2, raster=F) +
  NoLegend() + theme(aspect.ratio=1)
DimPlot(pbmc, reduction = reduction, cells.highlight = nn$index[2,], sizes.highlight = 2, raster=F) +
  NoLegend() + theme(aspect.ratio=1)

maxrow <- which(nn$distance == max(nn$distance), arr.ind = TRUE)[1]
nn$distance[maxrow,]

DimPlot(pbmc, reduction = reduction, cells.highlight = nn$index[maxrow,], sizes.highlight = 2, raster=F) +
  NoLegend() + theme(aspect.ratio=1)
```


Smooth UCell scores by averaging neighbors
```{r}
sig.cols <- names(signatures)
data <- pbmc

w.df <- vapply(sig.cols, FUN.VALUE = numeric(ncol(data)), FUN=function(s) {
  
  score.sig <- data@meta.data[,s]
  weighted.scores <- vapply(X = 1:nrow(nn$index),
                            FUN.VALUE = numeric(1),
                            FUN = function(x) {
                              r <- nn$index[x,]
                              r <- c(x,r)
                              
                              d <- nn$distance[x,]
                              d <- c(d[1],d)
                              
                              w <- 1/(0.01+d)
                              
                              sum(w * score.sig[r])/sum(w)
                            })
})
rownames(w.df) <- colnames(data)
colnames(w.df) <- paste0(colnames(w.df),"_NN")
head(w.df)
```

Add to metadata
```{r fig.height=7}
pbmc <- AddMetaData(pbmc, metadata = as.data.frame(w.df))

FeaturePlot(pbmc, reduction = reduction,
            features = c("CD4T","CD4T_NN","CD8T","CD8T_NN")) & theme(aspect.ratio = 1)

FeaturePlot(pbmc, reduction = reduction,
            features = c("Tcell_gd","Tcell_gd_NN","MAIT","MAIT_NN"))  & theme(aspect.ratio = 1)

FeaturePlot(pbmc, reduction = reduction,
            features = c("Treg","Treg_NN","NK","NK_NN"))  & theme(aspect.ratio = 1)
```

Smooth UCell scores by neighbor rank with decay function
```{r}
knn_smooth_scores <- function(
    matrix=NULL,
    nn=NULL,
    decay=0.2,   #decay must be bound between 0 and 1
    up.only=FALSE #scores can only increase
) {
  
  sig.cols <- colnames(matrix)
  w.df <- vapply(sig.cols, FUN.VALUE=numeric(nrow(matrix)), FUN=function(s) {
    
    ss.scores <- matrix[,s]
    
    weighted.scores <- vapply(X = seq_len(nrow(nn$index)),
                              FUN.VALUE = numeric(1),
                              FUN = function(x) {
                                r <- nn$index[x,]
                                r <- c(x,r)
                                
                                i <- seq(0, length(r)-1)
                                
                                w <- (1-decay)**i
                                
                                sum(w * ss.scores[r])/sum(w)
                              })
    if (up.only) {
      pmax(weighted.scores, ss.scores)
    } else {
      weighted.scores
    }
  })
  
  
  rownames(w.df) <- rownames(matrix)
  as.data.frame(w.df)
}  
```

```{r fig.height=7}
matrix <- pbmc@meta.data[,names(signatures)]

m <- knn_smooth_scores(matrix=matrix, nn=nn, decay=0.2)
colnames(m) <- paste0(colnames(m),"_NN")

pbmc <- AddMetaData(pbmc, metadata = m)

FeaturePlot(pbmc, reduction = reduction,
            features = c("CD4T","CD4T_NN","CD8T","CD8T_NN")) & theme(aspect.ratio = 1)

FeaturePlot(pbmc, reduction = reduction,
            features = c("Tcell_gd","Tcell_gd_NN","MAIT","MAIT_NN"))  & theme(aspect.ratio = 1)

FeaturePlot(pbmc, reduction = reduction,
            features = c("Treg","Treg_NN","NK","NK_NN"))  & theme(aspect.ratio = 1)
```

```{r fig.height=7}
matrix <- pbmc@meta.data[,names(signatures)]

m <- knn_smooth_scores(matrix=matrix, nn=nn, decay=1)
colnames(m) <- paste0(colnames(m),"_NN")
pbmc <- AddMetaData(pbmc, metadata = m)

FeaturePlot(pbmc, reduction = reduction,
            features = c("CD4T","CD4T_NN","CD8T","CD8T_NN")) & theme(aspect.ratio = 1)

FeaturePlot(pbmc, reduction = reduction,
            features = c("Tcell_gd","Tcell_gd_NN","MAIT","MAIT_NN"))  & theme(aspect.ratio = 1)

FeaturePlot(pbmc, reduction = reduction,
            features = c("Treg","Treg_NN","NK","NK_NN"))  & theme(aspect.ratio = 1)
```


```{r fig.height=7}
matrix <- pbmc@meta.data[,names(signatures)]

m <- knn_smooth_scores(matrix=matrix, nn=nn, decay=0.1, up.only = T)
colnames(m) <- paste0(colnames(m),"_NN")

pbmc <- AddMetaData(pbmc, metadata = m)

FeaturePlot(pbmc, reduction = reduction,
            features = c("CD4T","CD4T_NN","CD8T","CD8T_NN")) & theme(aspect.ratio = 1)

FeaturePlot(pbmc, reduction = reduction,
            features = c("Tcell_gd","Tcell_gd_NN","MAIT","MAIT_NN"))  & theme(aspect.ratio = 1)

FeaturePlot(pbmc, reduction = reduction,
            features = c("Treg","Treg_NN","NK","NK_NN"))  & theme(aspect.ratio = 1)
```




Using the new function implemented in UCell
```{r fig.height=6}
pbmc <- SmoothKNN(pbmc, signature.names = names(signatures))

FeaturePlot(pbmc, reduction = reduction,
            features = c("Treg","Treg_kNN","CD4T","CD4T_kNN")) & theme(aspect.ratio = 1)

pbmc <- SmoothKNN(pbmc, signature.names = names(signatures), decay=0.01, k = 50)
FeaturePlot(pbmc, reduction = reduction,
            features = c("Treg","Treg_kNN","CD4T","CD4T_kNN")) & theme(aspect.ratio = 1)

pbmc <- SmoothKNN(pbmc, signature.names = names(signatures), up.only = TRUE)
FeaturePlot(pbmc, reduction = reduction,
            features = c("Treg","Treg_kNN","CD4T","CD4T_kNN")) & theme(aspect.ratio = 1)


pbmc <- SmoothKNN(pbmc, signature.names = names(signatures), reduction="wnn.umap", suffix = "_UM")

FeaturePlot(pbmc, reduction = reduction,
            features = c("Treg","Treg_UM","CD4T","CD4T_UM")) & theme(aspect.ratio = 1)

```
Can we use it to smooth gene expression?
```{r fig.height=6}
pbmc <- SmoothKNN(pbmc, signature.names = c("CD4","CD8A"),
                  assay="RNA", reduction="pca", k=20)

a <- FeaturePlot(pbmc, reduction = reduction,
            features = c("CD4","CD8A"))

DefaultAssay(pbmc) <- "RNA_kNN"
b <- FeaturePlot(pbmc, reduction = reduction,
            features = c("CD4","CD8A"))
DefaultAssay(pbmc) <- "RNA"

a / b
```

With SingleCellExperiment
```{r fig.height=6}
library(SingleCellExperiment)
library(scater)
library(scran)
library(patchwork)
set.seed(123)

sce <- as.SingleCellExperiment(pbmc)
#clean up metadata
colData(sce) <- colData(sce)[,1:5]

sce <- ScoreSignatures_UCell(sce, features=signatures,
                             assay = "logcounts", name=NULL)

#PCA and UMAP
hvg <- getTopHVGs(sce, n=1000)
sce <- runPCA(sce, scale=TRUE, ncomponents=20, subset_row=hvg)
set.seed(1234)
sce <- runUMAP(sce, dimred="PCA")

pll <- lapply(names(signatures), function(x) {
    plotUMAP(sce, colour_by = x, by_exprs_values = "UCell") + 
       theme(aspect.ratio = 1)
})
wrap_plots(pll)
```

Signature smoothing
```{r fig.height=6}
sce <- SmoothKNN(sce, signature.names = names(signatures), 
                 sce.expname = "UCell", reduction = 'PCA')

pll <- lapply(paste0(names(signatures),"_kNN"), function(x) {
    plotUMAP(sce, colour_by = x, by_exprs_values = "UCell_kNN") + 
       theme(aspect.ratio = 1)
})
wrap_plots(pll)
```

Gene expression smoothing
```{r fig.height=4}
sce <- SmoothKNN(sce, signature.names = c("CD4","CD8A","TCF7"),
                 suffix = "_kNN", reduction = 'PCA',
                 sce.expname = "main", sce.assay = "logcounts")

altExp(sce, "main_kNN")
a <- plotUMAP(sce, colour_by = "CD8A", by_exprs_values = "logcounts")
b <- plotUMAP(sce, colour_by = "CD8A_kNN", by_exprs_values = "main_kNN")
a | b
```

Parallel processing
```{r}
library(BiocParallel)
t1 <- Sys.time()
pbmc <- SmoothKNN(pbmc, signature.names = names(signatures), reduction="pca")
t2 <- Sys.time()
t2 - t1

t1 <- Sys.time()
pbmc <- SmoothKNN(pbmc, signature.names = names(signatures), reduction="pca", BPPARAM = MulticoreParam(workers = 4))
t2 <- Sys.time()
t2 - t1
```

#Break cases:
1) k > number of cells
```{r}
cells <- colnames(pbmc)[1:5]
sub <- subset(pbmc, cells=cells)

sub <- SmoothKNN(sub, signature.names = names(signatures), reduction="pca")
```

2) number of cells==1
```{r}
cells <- colnames(pbmc)[1]
sub2 <- subset(pbmc, cells=cells)

sub2 <- SmoothKNN(sub2, signature.names = names(signatures), reduction="pca")
```

3) UCell scoring with 1 cell in object
```{r}
cells <- colnames(pbmc)[1]
sub2 <- subset(pbmc, cells=cells)

sub2 <- AddModuleScore_UCell(sub2, features = signatures)
```

4) k=0 or k=1
```{r}
tmp <- SmoothKNN(pbmc, signature.names = names(signatures), k=0)
tmp <- SmoothKNN(pbmc, signature.names = names(signatures), k=1)
```


