---
title: UCell demo
author: 
- Massimo Andreatta^[massimo.andreatta@unil.ch]
- Santiago Carmona^[santiago.carmona@unil.ch]
date: "15/02/2021"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
#output: html_notebook
---


Checks to do:

* Reproducibility, changing seeds, chunk size, ncores
* Accept matrix/data.frame with gene names as row names, convertible into data.table; sce object
* check presence of features and inform % missing and that all absent will be considered as 0



```{r message=F, warning=F}
#renv::restore()
library(ggplot2)
library(plotly)
#remotes::install_github("mojaveazure/seurat-disk")
library(SeuratDisk)
library(Seurat)
#remotes::install_git("https://gitlab.unil.ch/carmona/UCell.git")
#install.packages(pkgs="/Users/admin/gitlab/UCell_0.1.0.tgz",repos=NULL)

library(UCell)
library(BiocManager)
library(AUCell)
```

#Object at: https://drive.switch.ch/index.php/s/3kM5PQ0tQaG6d6A
```{r}
pbmc.Tcell <- readRDS("data/pbmc_multimodal.downsampled20k.Tcell.seurat.RNA.rds")
DimPlot(object = pbmc.Tcell, reduction = "umap", group.by = "celltype.l2", label = TRUE, label.size = 3, repel = TRUE) + NoLegend()
```

Define human T cell level 2 signatues
```{r, eval=F}
HCA.markers.Hs.Tcell <- list()
HCA.markers.Hs.Tcell$Tcell_CD4 <- c("CD4","CD40LG")
HCA.markers.Hs.Tcell$Tcell_CD8 <- c("CD8A","CD8B")
HCA.markers.Hs.Tcell$Tcell_Treg <- c("FOXP3")
HCA.markers.Hs.Tcell$Tcell_MAIT <- c("KLRB1", "SLC4A10", "NCR3")
HCA.markers.Hs.Tcell$Tcell_gd <- c("TRDC", "TRGC1", "TRGC2", "TRDV1")
HCA.markers.Hs.Tcell$Tcell_NK <- c("FGFBP2", "SPON2", "KLRF1", "FCGR3A", "KLRD1", "TRDC")
saveRDS(HCA.markers.Hs.Tcell,"data/HCA.markers.Hs.Tcell.PBMC.RData")
```



```{r}
Idents(pbmc.Tcell) <- "celltype.l1"
pbmc.Tcell.CD8 <- subset(pbmc.Tcell, idents = c("CD8 T"))
table(pbmc.Tcell.CD8@active.ident)
DimPlot(object = pbmc.Tcell.CD8, reduction = "umap", group.by = "celltype.l2", label = TRUE, label.size = 3, repel = TRUE) + NoLegend()
```


```{r}
HCA.markers.Hs.Tcell <- readRDS("data/HCA.markers.Hs.Tcell.PBMC.RData")
pbmc.Tcell.CD8 <- AddModuleScore(pbmc.Tcell.CD8, features = list(Tcell_CD8=c("CD8A","CD8B")), name="Tcell_CD8_test", seed=123)
VlnPlot(pbmc.Tcell.CD8, features = "Tcell_CD8_test1")

pbmc.Tcell <- AddModuleScore(pbmc.Tcell, features = list(Tcell_CD8=c("CD8A","CD8B")), name="Tcell_CD8_test", seed=123)
VlnPlot(pbmc.Tcell, features = "Tcell_CD8_test1")
```



#check if rank matrix is present. if not, re-calculate
# store rank matrix in @misc

#show performance for 1st time and for 2nd (precalulacted ranks) time

```{r}
pbmc.Tcell.CD8 <- AddModuleScore_UCell(pbmc.Tcell.CD8, features = list(Tcell_CD8=c("CD8A","CD8B")))
VlnPlot(pbmc.Tcell.CD8, features = "Tcell_CD8_UCell")

pbmc.Tcell <- AddModuleScore_UCell(pbmc.Tcell, features = list(Tcell_CD8=c("CD8A","CD8B")))
VlnPlot(pbmc.Tcell, features = "Tcell_CD8_UCell")
```



```{r}
plot(pbmc.Tcell.CD8$Tcell_CD8_test1,pbmc.Tcell.CD8$Tcell_CD8_UCell)
plot(pbmc.Tcell$Tcell_CD8_test1,pbmc.Tcell$Tcell_CD8_UCell)

```


Compare to AUCell



```{r}
cells_rankings <- AUCell_buildRankings(pbmc.Tcell@assays$RNA@data,nCores = 1, plotStats = F)
cells_AUC <- AUCell_calcAUC(HCA.markers.Hs.Tcell, cells_rankings, aucMaxRank=1000)
cells_AUC.num <- as.data.frame(t(getAUC(cells_AUC)))
pbmc.Tcell <- AddMetaData(pbmc.Tcell,cells_AUC.num$Tcell_CD8,col.name = "Tcell_CD8_AUCell")
pbmc.Tcell <- AddMetaData(pbmc.Tcell,cells_AUC.num)
VlnPlot(pbmc.Tcell, features = "Tcell_CD8_AUCell")
VlnPlot(pbmc.Tcell, features = "Tcell_CD8")
```


```{r}
cells_rankings2 <- AUCell_buildRankings(pbmc.Tcell.CD8@assays$RNA@data,nCores = 1, plotStats = T)
cells_AUC2 <- AUCell_calcAUC(HCA.markers.Hs.Tcell, cells_rankings2, aucMaxRank=1000)
cells_AUC.num2 <- as.data.frame(t(getAUC(cells_AUC2)))
pbmc.Tcell.CD8 <- AddMetaData(pbmc.Tcell.CD8,cells_AUC.num2$Tcell_CD8,col.name = "Tcell_CD8_AUCell")
VlnPlot(pbmc.Tcell.CD8, features = "Tcell_CD8_AUCell")
```



```{r}
plot(pbmc.Tcell.CD8$Tcell_CD8_test1,pbmc.Tcell.CD8$Tcell_CD8_UCell)
plot(pbmc.Tcell.CD8$Tcell_CD8_AUCell,pbmc.Tcell.CD8$Tcell_CD8_UCell)
```



Test ranks storing functionality

```{r}
pbmc.Tcell.UcellPrecomp <- AddModuleScore_UCell(pbmc.Tcell, features = list(Tcell_CD8=c("CD8A","CD8B")), storeRanks = T)
VlnPlot(pbmc.Tcell.UcellPrecomp, features = "Tcell_CD8_UCell")
```

```{r}
pbmc.Tcell.UcellPrecomp <- AddModuleScore_UCell(pbmc.Tcell.UcellPrecomp, features = HCA.markers.Hs.Tcell)
VlnPlot(pbmc.Tcell.UcellPrecomp, features = paste0(names(HCA.markers.Hs.Tcell),"_UCell"))

```

Gene ranks are stored here:
```{r}
head(pbmc.Tcell.UcellPrecomp@misc$UCell$RNA$cells_rankings[,1:10])
head(pbmc.Tcell.UcellPrecomp@misc$UCell$RNA$cells_rankings["CD8A",1:10])
hist(as.numeric(pbmc.Tcell.UcellPrecomp@misc$UCell$RNA$cells_rankings["CD8A",]))

#head(pbmc.Tcell.UcellPrecomp@misc$UCell$RNA$cells_rankings["CD8A",1:10,on="rn"])
#hist(as.numeric(pbmc.Tcell.UcellPrecomp@misc$UCell$RNA$cells_rankings["CD8A",-1,on="rn"]))
```




Small benchmark
```{r}
testSamp <- c(1e3,2e3,3e3,4e3)
#testSamp <- c(1e3,5e3,1e4)

features <- HCA.markers.Hs.Tcell
obj <- pbmc.Tcell
  
time_table <- matrix(NA,nrow = length(testSamp),ncol = 3)
colnames(time_table) <- c("size","AUCell","UCell")

memory_table <- matrix(NA,nrow = length(testSamp),ncol = 3)
colnames(memory_table) <- c("size","AUCell","UCell")



for (i in seq_along(testSamp)[]){
  i.samp <- testSamp[i]
  time_table[i,1] <- i.samp
  memory_table[i,1] <- i.samp

  print(i.samp)
  Rprof(tf <- "rprof.log", memory.profiling=TRUE)
  t <- system.time({
    cells_rankings <- AUCell_buildRankings(obj@assays$RNA@data[,1:i.samp],nCores = 1, plotStats = F)
    cells_AUC <- AUCell_calcAUC(features, cells_rankings, aucMaxRank=1000)
    cells_AUC.num <- as.data.frame(t(getAUC(cells_AUC)))
    obj_out1 <- AddMetaData(obj,cells_AUC.num)
    })
  Rprof(NULL)
  
  memPeak <- max(summaryRprof("Rprof.log", memory="both")$by.total$mem.total)
  time_table[i,2] <- t[["elapsed"]]
  memory_table[i,2] <- memPeak
  
  print(obj_out1@meta.data[1:5,])
  
  Rprof(tf <- "rprof.log", memory.profiling=TRUE)
  t <- system.time({
    obj_out2 <- AddModuleScore_UCell(obj[,1:i.samp], features = features, maxRank = 1000)
  })
  Rprof(NULL)
  
  memPeak <- max(summaryRprof("Rprof.log", memory="both")$by.total$mem.total)  
  time_table[i,3] <- t[["elapsed"]]
  memory_table[i,3] <- memPeak
  
#  print(obj_out1@meta.data[1:5,])
  
}

```

```{r}

library(tidyr)

time_table.df <- time_table %>% as.data.frame() %>% pivot_longer(-size,values_to="time") 

ggplot(time_table.df, aes(x=size, y = time, name)) +  geom_point(aes(color = name))

memory_table.df <- memory_table %>% as.data.frame() %>% pivot_longer(-size,values_to="memory") 

ggplot(memory_table.df, aes(x=size, y = memory, name)) +  geom_point(aes(color = name)) + ylim(c(0,NA))




```




