---
title: UCell with signatures from mSigDB
author: 
- Massimo Andreatta^[massimo.andreatta@unil.ch]
- Santiago Carmona^[santiago.carmona@unil.ch]
date: "15/02/2021"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
#output: html_notebook
---


```{r message=F, warning=F}
#renv::restore()
library(Seurat)
#Development moved to GitHub
remotes::install_github("carmonalab/UCell")

library(UCell)
```

The [original dataset](https://atlas.fredhutch.org/data/nygc/multimodal/pbmc_multimodal.h5seurat) is very large (>160K cells), for this illustrative example we used a downsampled version (20,000 cells)
```{r, eval=T}
set.seed(12345)
library(SeuratDisk)
pbmc.azimuth <- LoadH5Seurat("pbmc_multimodal.h5seurat")
pbmc.azimuth <-  subset(pbmc.azimuth, cells = sample(Cells(pbmc.azimuth), 20000))

Idents(pbmc.azimuth) <- "celltype.l1"

saveRDS(pbmc.azimuth,"pbmc_multimodal.downsampled20k.seurat.RNA.rds")
```

Have a look at the atlas in UMAP space
```{r}
library(ggplot2)
DimPlot(object = pbmc.azimuth, reduction = "wnn.umap", group.by = "celltype.l1", label = TRUE, label.size = 3, repel = TRUE) +
    ggtitle("Level 1 annotations")
DimPlot(object = pbmc.azimuth, reduction = "wnn.umap", group.by = "celltype.l2", label = TRUE, label.size = 3, repel = TRUE) +
    ggtitle("Level 2 annotations") + NoLegend()

```


## Score signatures using UCell

Get some signatures from MSigDB
```{r}
#install.packages("msigdbr")
library(msigdbr)
h_gene_sets = as.data.frame(msigdbr(species = "Homo sapiens", category = "H"))

names <- unique(h_gene_sets$gs_name)
h.markers <- list()

for (h in names) {
   sub <- h_gene_sets[h_gene_sets$gs_name==h,"gene_symbol"]
   h.markers[[h]] <- unique(sub)
}

length(h.markers)
head(h.markers)
```

Then run AddModuleScore_UCell to calculate signature directly from the Seurat object, and store results in the object metadata.
```{r}
pbmc.azimuth <- AddModuleScore_UCell(pbmc.azimuth, features = h.markers, ncores=8)
signature.names <- paste0(names(h.markers),"_UCell")
```

```{r fig.height=5}
VlnPlot(pbmc.azimuth, features = signature.names[1:9], group.by = "celltype.l1", pt.size = 0)
VlnPlot(pbmc.azimuth, features = signature.names[10:18], group.by = "celltype.l1", pt.size = 0)
VlnPlot(pbmc.azimuth, features = signature.names[19:27], group.by = "celltype.l1", pt.size = 0)
VlnPlot(pbmc.azimuth, features = signature.names[28:36], group.by = "celltype.l1", pt.size = 0)
VlnPlot(pbmc.azimuth, features = signature.names[37:45], group.by = "celltype.l1", pt.size = 0)
VlnPlot(pbmc.azimuth, features = signature.names[45:50], group.by = "celltype.l1", pt.size = 0)
```

#Read in signatures from Human Cell atlas [ref]
```{r}
tab <- read.csv(file="aux/HCL1_selected_signatures.csv", header=T)
head(tab)

names <- colnames(tab)
cd45.markers <- list()

for (h in names) {
   cd45.markers[[h]] <- unique(tab[,h])
}

head(cd45.markers$B.cell)

```

Then run AddModuleScore_UCell to calculate signature directly from the Seurat object, and store results in the object metadata.
```{r}
pbmc.azimuth <- AddModuleScore_UCell(pbmc.azimuth, features = cd45.markers, ncores=8)
signature.names <- paste0(names(cd45.markers),"_UCell")
```

```{r fig.height=4}
VlnPlot(pbmc.azimuth, features = signature.names, group.by = "celltype.l1", pt.size = 0)
VlnPlot(pbmc.azimuth, features = signature.names, group.by = "celltype.l2", pt.size = 0)
```

See in UMAP space
```{r fig.height=3}
FeaturePlot(pbmc.azimuth, reduction = "wnn.umap", features = signature.names, ncol=3, order=T, keep.scale="all")
```

What is the influence of signature length? down-size signatures to observe effect
```{r}
set.seed(123)
length(cd45.markers$B.cell)
cd45.markers$B.cell.100 <- sample(cd45.markers$B.cell, size = 100)
cd45.markers$B.cell.50 <- sample(cd45.markers$B.cell, size = 50)
cd45.markers$B.cell.20 <- sample(cd45.markers$B.cell, size = 20)
cd45.markers$B.cell.10 <- sample(cd45.markers$B.cell, size = 10)
set.seed(1234)
cd45.markers$B.cell.100.2 <- sample(cd45.markers$B.cell, size = 100)
cd45.markers$B.cell.50.2 <- sample(cd45.markers$B.cell, size = 50)
cd45.markers$B.cell.20.2 <- sample(cd45.markers$B.cell, size = 20)
cd45.markers$B.cell.10.2 <- sample(cd45.markers$B.cell, size = 10)
set.seed(12345)
cd45.markers$B.cell.100.3 <- sample(cd45.markers$B.cell, size = 100)
cd45.markers$B.cell.50.3 <- sample(cd45.markers$B.cell, size = 50)
cd45.markers$B.cell.20.3 <- sample(cd45.markers$B.cell, size = 20)
cd45.markers$B.cell.10.3 <- sample(cd45.markers$B.cell, size = 10)
```

```{r fig.height=6}
pbmc.azimuth <- AddModuleScore_UCell(pbmc.azimuth, features = cd45.markers, ncores=8)
signature.names <- paste0(names(cd45.markers),"_UCell")

VlnPlot(pbmc.azimuth, features = signature.names, group.by = "celltype.l1", pt.size = 0)
```

```{r fig.height=4}
b.cells <- subset(pbmc.azimuth, subset=celltype.l1=="B")
VlnPlot(b.cells, features = signature.names, group.by = "celltype.l1", pt.size = 0, same.y.lims = T, ncol=6)

```


