---
title: Evaluating human TIL subtype signatures using UCell
author: "M. Andreatta and S. Carmona"
date: "02/02/2022"
output:
  rmdformats::readthedown:
    self-contained: true
    highlight: haddock
    thumbnails: false
    css: styles.css
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file, encoding=encoding,
  output_file=file.path(dirname(input_file), out_dir, 'UCell_vignette_TILstates.html'))})
---



In this vignette we show how to use [UCell](https://github.com/carmonalab/UCell) gene signature scoring to identify and filter tumor-infiltrating T cell subtypes


```{r setup, echo=FALSE, message=F, warning=F, results=F}
#install.packages("rmdformats")
#Template markdown setup
library(knitr)
library(rmdformats)
library(formatR)


## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               cache.lazy=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               dev='png')
opts_knit$set(width=75)

```

## Prepare environment

```{r message=F, warning=F, results=F}
remotes::install_github("carmonalab/UCell", ref="v1.3")

library(Seurat)
library(UCell)
library(dplyr)
set.seed(123)
```

## Query single-cell data

Use data from [Yost et al. Nat Med 2019](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6689255/)
Pre-compiled R object available [here](https://drive.switch.ch/index.php/s/cluBLHkFFzLZWzL/download)
```{r fig.height=2.5}
inputFile <- "Yost.pretreatment.all.rds"
if(!file.exists(inputFile)) { download.file("https://drive.switch.ch/index.php/s/cluBLHkFFzLZWzL/download",inputFile)}
data.seurat <- readRDS(inputFile)
data.seurat <- data.seurat %>% NormalizeData %>% FindVariableFeatures %>% ScaleData %>% RunPCA %>% RunUMAP(dims=1:30)
data.seurat
```

Colour by patient and original cluster annotations

```{r}
DimPlot(data.seurat, group.by = "patient")
DimPlot(data.seurat, group.by = "cluster")
DimPlot(data.seurat, group.by = "cluster.T")
```

## Unsupervised clustering

```{r}
set.seed(123)

which.red <- "pca"
resol=0.7
ndim=30

data.seurat <- FindNeighbors(data.seurat, reduction = which.red, dims = 1:ndim)
data.seurat <- FindClusters(data.seurat, resolution = resol)

DimPlot(data.seurat, reduction="umap", group.by = "seurat_clusters", label = T) + NoLegend()
```

## Score general cell type signatures using UCell

Apply UCell with human cell type signatures to identify major cell types
These signatures were extracted from [Han et al Nature 2020](https://www.nature.com/articles/s41586-020-2157-4) and further filtered

```{r, fig.height=12, fig.width=7}

signaturesHumanCellTypes <- readRDS("aux/signaturesHumanCellTypes.rds")

head(signaturesHumanCellTypes)

data.seurat <- AddModuleScore_UCell(data.seurat, features=signaturesHumanCellTypes, ncores=4)

#Some major cell types to look at:
toplot <- c("Macrophage","Fibroblast","T.cell",
            "Stromal.cell","B.cell","Myeloid.cell",
            "Endothelial.cell.1","NK")

featnames <- paste0(toplot,"_UCell")
FeaturePlot(data.seurat, features = featnames, pt.size=0.1, max.cutoff = 'q99', ncol=2)
VlnPlot(data.seurat, features = featnames, pt.size = 0, split.by = "seurat_clusters", ncol=2)
```

We can appreciate that some clusters have a clearly high mean distribution of T cell signature scores (T.cell_UCell score > 0.2).
We can use this value as a rudimentary threshold to split T cell from non-T cell clusters.


## Identify T cells based on signatures and subset them

Identify T cell clusters by UCell score

```{r}
# select as Tcell clusters only those with median Uscore>0.2
medians <- sapply(levels(data.seurat$seurat_clusters), function(x){
   median(data.seurat@meta.data[data.seurat$seurat_clusters==x, "T.cell_UCell"])
})
tcell.clusters <- names(medians[medians>0.2])

#Add metadata
data.seurat$is.Tcell <- FALSE
data.seurat@meta.data[data.seurat$seurat_clusters %in% tcell.clusters, "is.Tcell"] <- TRUE
DimPlot(data.seurat, group.by = "is.Tcell")
```

Subset on T cells

```{r}
data.seurat.tcells <- subset(data.seurat, subset=is.Tcell==TRUE)
data.seurat.tcells
```

Recalculate embeddings only for filtered T cells

```{r}
ndim=20
resol=3
data.seurat.tcells <- data.seurat.tcells %>% NormalizeData %>% FindVariableFeatures %>% ScaleData %>% RunPCA  %>% RunUMAP(dims=1:ndim)
```

By patient and by annotation from original study

```{r}
DimPlot(data.seurat.tcells, group.by = "patient")
DimPlot(data.seurat.tcells, group.by = "cluster.T")
DimPlot(data.seurat.tcells, group.by = "seurat_clusters")
```

## Score TIL subtype-specific signatures using UCell

Now we can apply UCell using signatures specific for distict T cell subtypes.
TIL states signatures were obtained from ProjecTILs' reference TIL atlas (https://github.com/carmonalab/ProjecTILs), Andreatta et al Nature Communications (2021)

```{r}
signaturesHumanTILs <- readRDS("aux/signaturesHumanTILs.rds")
signaturesHumanTILs
signaturesHumanTILs[["cycling"]] <- c("TOP2A","MKI67","STMN1")

data.seurat.tcells <- AddModuleScore_UCell(data.seurat.tcells, features=signaturesHumanTILs, ncores=4)
```

```{r}
featnames <- paste0(names(signaturesHumanTILs),"_UCell")
FeaturePlot(data.seurat.tcells, features = featnames, pt.size=0.1,order = T)
VlnPlot(data.seurat.tcells, features = featnames, pt.size = 0, split.by = "seurat_clusters")
```


## Compare TIL subtype signature scores against original annotation

Now we can assess gene signature scores (their average values) in each T cell cluster as defined by Yost et al. 
An overall agreement between T cell subtypes defined by Yost and the corresponding gene signature scores can be observed.For example, CD8_eff and CD8_mem both with high CD8_EffectorMemory score, CD8_ex_act and CD_ex with highest CD8_Tex score, Treg with highest Treg score and Th17 with highest Thelper score. Instead, other Yost clusters seem to be more heterogeneous/ambiguously defined, e.g. CD8_act with both high CD8_EffectorMemory and CD4 Thelper scores, and Naive cluster with high Thelper signal.


```{r}
library(dplyr)
TILsigScores <- data.seurat.tcells@meta.data[,featnames]
TILsigScores_vs_OriginalCluster <- TILsigScores %>% filter(!is.na(data.seurat.tcells@meta.data$cluster.T)) %>% group_by(data.seurat.tcells@meta.data$cluster.T[!is.na(data.seurat.tcells@meta.data$cluster.T)]) %>% summarise_each(mean)
TILsigScores_vs_OriginalCluster.m <- as.matrix(TILsigScores_vs_OriginalCluster[,-1])
rownames(TILsigScores_vs_OriginalCluster.m) <- TILsigScores_vs_OriginalCluster[,1][[1]]
heatmap(t(TILsigScores_vs_OriginalCluster.m),cexCol=0.7,scale="none")
```


For more in-depth analysis of T cell states we recommend projecting your data onto T cell reference atlases using [ProjecTILs](https://github.com/carmonalab/ProjecTILs)

## Further reading

For more examples of UCell functionalities and reproducible R Notebook see [UCell Demos](https://gitlab.unil.ch/carmona/UCell_demo/)

The code and the package are available at the UCell [GitHub repository](https://github.com/carmonalab/UCell)