---
title: Comparing UCell outputs from R and python
author: 
- Massimo Andreatta^[massimo.andreatta@unige.ch]
- Santiago Carmona^[santiago.carmona@unige.ch]
date: "08/01/2026"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'UCell_anndata.html'))})
#output: html_notebook
---

This notebook reproduces results from pyUCell with the R version of UCell. We will use a test object from scanpy, also used in pyUCell's demo.

```{r message=F, warning=F}
library(UCell)
library(reticulate)
library(zellkonverter)
library(Seurat)
library(ggplot2)
```

Download anndata object from scanpy. This requires "reticulate" to run python from within R. Alternatively, you may download and store the `pbmc3k.h5ad` from a python session.
```{r}
do_download <- FALSE

if (do_download) {
  use_python("/usr/bin/python3", required = TRUE)
  sc <- import("scanpy")
  
  adata <- sc$datasets$pbmc3k()
  adata$write("pbmc3k.h5ad")
}
```

Read object into R. For conversion, we will use `zellkoverter`; first converts an ".h5ad" object to "sce" format, then we can convert from "sce" to "Seurat". Conversions between R and python objects are known to be clunky.
```{r}
sce <- readH5AD("pbmc3k.h5ad")
pbmc <- as.Seurat(sce, counts="X", data=NULL)

pbmc
```
Define two simple signatures.
```{r}
signatures = list()
signatures[['T_cell']] <- c('CD3D', 'CD3E', 'CD2')
signatures[['B_cell']] <- c('MS4A1', 'CD79A', 'CD79B')
```

Run UCell on the PBMC object to evaluate the T cell and B cell signatures.
```{r}
pbmc <- AddModuleScore_UCell(pbmc, features = signatures)
```

Have a look at the results in metadata.
```{r}
head(pbmc[[]])
```

Calculate low-dimensional embeddings, to see the signature activity in UMAP space.
```{r}
pbmc <- pbmc |> NormalizeData() |> FindVariableFeatures(nfeatures = 500) |>
  ScaleData() |> RunPCA(npcs = 10) |> RunUMAP(dims = 1:10)
```

Plot UCell scores in UMAP space
```{r fig.width=10, fig.height=5}
FeaturePlot(pbmc, features=c("T_cell_UCell","B_cell_UCell")) & theme(aspect.ratio = 1)
```



