---
title: UCell signature enrichment in spatial data (CosMx)
author: "M. Andreatta and S. Carmona"
date: "08/01/2026"
output:
  rmdformats::readthedown:
    self-contained: true
    highlight: haddock
    thumbnails: false
    css: styles.css
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file, encoding=encoding,
  output_file=file.path(dirname(input_file), out_dir, 'UCell_cosmx.html'))})
---

```{r setup, echo=FALSE, message=F, warning=F, results=F}
#install.packages("rmdformats")
#Template markdown setup
library(knitr)
library(rmdformats)
library(formatR)
library(renv)
#renv::restore()

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               cache.lazy=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               dev='png')
opts_knit$set(width=75)

```

## Installation

```{r message=F, warning=F, results=F}
#if (!requireNamespace("BiocManager", quietly=TRUE))
#    install.packages("BiocManager")
#BiocManager::install("UCell")

#remotes::install_github("mojaveazure/seurat-disk")
#library(SeuratDisk)


library(Seurat)
library(SeuratObject)
library(ggplot2)
library(UCell)
set.seed(42)
```


Get object with pre-processed CosMx data from: https://zenodo.org/records/14330691
Direct download link: https://zenodo.org/records/14330691/files/CosMx_BCC_Yerly_2024.rds?download=1

NOTE: the object on Zenodo already contains meta-data after processing (e.g. predicted cell types, homotypic/heterotypic cancer cells, MP scores, etc.). Since no cells were removed, and transcript data were not subject to transformation, we can start analyses from this object, ignoring the meta-data.

```{r}
data_dir <- "./"
seu.obj <- readRDS(paste0(data_dir,"../Yerly_BCC_wounding/CosMx/CosMx_BCC_Yerly_2024.rds"))
```

```{r}
pal <- c("#F2E9E4","#C9ADA7","#9A8C98","#4A4E69")

seu.obj$Patient_ID <- factor(seu.obj$Patient_ID, levels=c("PatientA","PatientB","PatientC","PatientD"))

Idents(seu.obj) <- "Patient_ID"
VlnPlot(seu.obj, features="nFeature_Nanostring", pt.size = 0, log = TRUE, cols=pal)
VlnPlot(seu.obj, features="nFeature_Nanostring", pt.size = 0, log = FALSE, cols=pal)
ggsave("plots/cosmx_nfeatures.pdf", height=3.5, width=4.5)
```
Signatures
```{r}
sigs <- list()
sigs$Fibroblast <- c("COL1A1","COL1A2","COL3A1","DCN","LUM","THY1","KRT18-","KRT19-")
sigs$Keratinocyte <- c("KRT5","KRT14","KRT1","KRT10","KRTDAP")
```

Run UCell
```{r}
library(UCell)
seu.obj <- AddModuleScore_UCell(seu.obj, features=sigs, name="", ncores=8, maxRank = 400)
```


Plot MP activity for specific FOV

```{r}
# Set your desired image and FOV 

use_fov <- 45 # FOV desired 
use_slide_metadata <- unique(seu.obj@meta.data$Run_Tissue_name.x)[1] # Slide desired, as named in the metadata column ‘Run_Tissue_name’ 
use_slide_image <- Images(seu.obj)[1] # Slide desired, as named in images 

cells_of_interest <- seu.obj$id[(seu.obj$fov == use_fov) & ( seu.obj$Run_Tissue_name.x == use_slide_metadata)] 
centroid_data <- seu.obj@images[[use_slide_image]]$centroids 
zoom_fov <- apply(centroid_data@coords[centroid_data@cells %in% cells_of_interest,], 2, range)

cols = c("white", "darkred")

a <- ImageFeaturePlot(seu.obj,
                        fov = use_slide_image,
                        features = "Fibroblast",
                        border.color = "black",
                        cols = cols,
                        border.size = 0.05,
                        dark.background = F,
                        coord.fixed = T,
                        min.cutoff = 0, max.cutoff = 1) +
  xlim(zoom_fov[, 2]) + ylim(zoom_fov[, 1]) + coord_flip() + theme(aspect.ratio = 1)

# Keratinocytes

cols = c("white", "darkblue")

b <- ImageFeaturePlot(seu.obj,
                        fov = use_slide_image,
                        features = "Keratinocyte",
                        border.color = "black",
                        cols = cols,
                        border.size = 0.1,
                        dark.background = F,
                        coord.fixed = T,
                        min.cutoff = 0, max.cutoff = 1) +
  xlim(zoom_fov[, 2]) + ylim(zoom_fov[, 1]) + coord_flip() + theme(aspect.ratio = 1)


a | b

ggsave("plots/cosmx_sigs_slide1_fov45.pdf", height=4, width=9)
```

```{r}
# Set your desired image and FOV 

use_fov <- 15 # FOV desired 
use_slide_metadata <- unique(seu.obj@meta.data$Run_Tissue_name.x)[1] # Slide desired, as named in the metadata column ‘Run_Tissue_name’ 
use_slide_image <- Images(seu.obj)[1] # Slide desired, as named in images 

cells_of_interest <- seu.obj$id[(seu.obj$fov == use_fov) & ( seu.obj$Run_Tissue_name.x == use_slide_metadata)] 
centroid_data <- seu.obj@images[[use_slide_image]]$centroids 
zoom_fov <- apply(centroid_data@coords[centroid_data@cells %in% cells_of_interest,], 2, range)

cols = c("white", "darkred")

a <- ImageFeaturePlot(seu.obj,
                        fov = use_slide_image,
                        features = "Fibroblast",
                        border.color = "black",
                        cols = cols,
                        border.size = 0.05,
                        dark.background = F,
                        coord.fixed = T,
                        min.cutoff = 0, max.cutoff = 1) +
  xlim(zoom_fov[, 2]) + ylim(zoom_fov[, 1]) + coord_flip() + theme(aspect.ratio = 1)

# Keratinocytes

cols = c("white", "darkblue")

b <- ImageFeaturePlot(seu.obj,
                        fov = use_slide_image,
                        features = "Keratinocyte",
                        border.color = "black",
                        cols = cols,
                        border.size = 0.1,
                        dark.background = F,
                        coord.fixed = T,
                        min.cutoff = 0, max.cutoff = 1) +
  xlim(zoom_fov[, 2]) + ylim(zoom_fov[, 1]) + coord_flip() + theme(aspect.ratio = 1)


a | b

ggsave("plots/cosmx_sigs_slide1_fov15.pdf", height=4, width=9)
```

