---
title: UCell signature enrichment - interacting with Seurat
author: "M. Andreatta and S. Carmona"
date: "29/04/2025"
output:
  rmdformats::readthedown:
    self-contained: true
    highlight: haddock
    thumbnails: false
    css: styles.css
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file, encoding=encoding,
  output_file=file.path(dirname(input_file), out_dir, 'UCell_Seurat_vignette.html'))})
---

```{r setup, echo=FALSE, message=F, warning=F, results=F}
#install.packages("rmdformats")
#Template markdown setup
library(knitr)
library(rmdformats)
library(formatR)
library(renv)
#renv::restore()

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               cache.lazy=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               dev='png')
opts_knit$set(width=75)

```

## Installation

```{r message=F, warning=F, results=F}
#if (!requireNamespace("BiocManager", quietly=TRUE))
#    install.packages("BiocManager")
#BiocManager::install("UCell")

#remotes::install_github("mojaveazure/seurat-disk")
#library(SeuratDisk)


library(Seurat)
library(SeuratObject)
library(ggplot2)
library(UCell)
set.seed(42)
```

## Prepare data

From a large PBMC datasets, extract a few populations with clear markers. These will be used to illustrate signature scores.

Download reference from: https://atlas.fredhutch.org/data/nygc/multimodal/pbmc_multimodal.h5seurat
```{r}

#data <- LoadH5Seurat("data/pbmc_multimodal.h5seurat")
data <- readRDS("../BCC_Kuonen/cache/BCC_all_LY_clean_annotated.rds")
```

```{r}
DimPlot(data)
```
Helper function for plots
```{r}
simplify_plot <- function() {
  theme(
    axis.title = element_blank(),
    axis.text  = element_blank(),
    axis.ticks = element_blank(),
    aspect.ratio = 1
  )
}
```

Subset on RNA and on three populations
```{r}
data <- subset(data, annotation %in% c("Tcell","Melanocyte"))

data <- FindVariableFeatures(data, nfeatures = 1000)
data <- RunPCA(data, npcs = 20)
data <- RunUMAP(data, dims = 1:20)
```

```{r}
DimPlot(data, group.by = "annotation")
```

```{r}
soft_three_colors <- c(
  "#A6CEE3",
  "#B2DF8A",
  "#FDBF6F"   
)
```

```{r fig.height=4}
data <- FindNeighbors(data, reduction = "pca")
data <- FindClusters(data, resolution = 0.05)
DimPlot(data, group.by = "seurat_clusters") & simplify_plot() & scale_color_manual(values = soft_three_colors)

ggsave("plots/toy_clusters_umap.pdf", width = 4, height=3)
```

Make some signatures to highlight clusters
```{r}
mark <- FindAllMarkers(data, group.by = "seurat_clusters", only.pos = 1, max.cells.per.ident = 1000, min.diff.pct = 0.1)

head(mark[mark$cluster==0,],10)
head(mark[mark$cluster==1,],10)
head(mark[mark$cluster==2,],10)
```

```{r fig.height=5, fig.width=10}
FeaturePlot(data, features = c("CD4","CD8A","CD8B","CCL5","NKG7","RAB27A"), ncol=3) & theme(aspect.ratio=1)
```

```{r}
sigs <- list()
sigs$Tcell <- c("CD2","CD3E","CD3D","CD3G","TRAC")
sigs$cytotox <- c("NKG7","GZMA","PRF1","GLNY")
sigs$Tmem <- c("IL7R","TCF7","LTB","CCR7")
sigs$CD8T <- c("CD8A","CD8B")
sigs$CD4T <- c("CD4","CD40LG")
sigs$melanocyte <- c("PMEL","MLANA","CAPN3","TYR","RAB27A")

data <- AddModuleScore_UCell(data, features = sigs, name = "")
```

```{r fig.height=8, fig.width=14}
library(viridis)

p <- FeaturePlot(data, features = names(sigs), ncol=3, max.cutoff = "q99") & simplify_plot()
#p

p & scale_color_viridis_c(option = "H") & theme(legend.position = "right")

#custom_colors <- c(
#  "#D9D9D1",  # light gray
#  "#FFF176",  # soft yellow
#  "#FFB74D",  # light orange
#  "#FF7043",  # medium orange-red
#  "#E53935",  # red
#  "#8B0000"   # dark red
#)

custom_colors <- c(
  "#E0E0E0",  # light gray (lowest values)
  "#ADD8E6",  # light blue
  "#6495ED",  # medium blue
  "#1E90FF",  # strong blue
  "#0B3D91"   # dark blue (highest values)
)

my_palette <- colorRampPalette(custom_colors)

p & scale_color_gradientn(colors = my_palette(100))

```
Save plot
```{r}
p <- FeaturePlot(data, features = "melanocyte", max.cutoff = "q99") & simplify_plot()
p & scale_color_gradientn(colors = my_palette(100))


ggsave("plots/toy_melanocyte_umap.pdf", width = 5, height=4)
```

kNN smoothing
```{r fig.height=8, fig.width=14}
data <- SmoothKNN(data, signature.names = names(sigs), k = 20, decay = 0.1)

p <- FeaturePlot(data, features = paste0(names(sigs), "_kNN"), ncol=3, max.cutoff = "q99") & simplify_plot()
p & scale_color_gradientn(colors = my_palette(100))
```

```{r fig.height=8, fig.width=14}
data <- SmoothKNN(data, signature.names = names(sigs), k = 20, decay = 0.05)

p <- FeaturePlot(data, features = paste0(names(sigs), "_kNN"), ncol=3, max.cutoff = "q99") & simplify_plot()
p & scale_color_gradientn(colors = my_palette(100))
```


```{r}
p <- FeaturePlot(data, features = c("cytotox","cytotox_kNN"), ncol=2, max.cutoff = "q99") & simplify_plot()
p & scale_color_gradientn(colors = my_palette(100))

p <- FeaturePlot(data, features = c("CD4T","CD4T_kNN"), ncol=2, max.cutoff = "q99") & simplify_plot()
p & scale_color_gradientn(colors = my_palette(100))

ggsave("plots/toy_cd4_smooth_umap.pdf", width = 10, height=4)

p <- FeaturePlot(data, features = c("melanocyte","melanocyte_kNN"), ncol=2, max.cutoff = "q99") & simplify_plot()
p & scale_color_gradientn(colors = my_palette(100))

```

```{r fig.height=2.5, fig.width=3}
VlnPlot(data, features = c("cytotox_kNN"), group.by = "seurat_clusters", pt.size=0, cols = soft_three_colors) 
VlnPlot(data, features = c("melanocyte_kNN"), group.by = "seurat_clusters", pt.size=0, cols = soft_three_colors)

ggsave("plots/melanocyte_violin_knn.pdf", height=2.5, width=2.5)
```

```{r fig.height=2.5, fig.width=3}
max.u <- max(data$CD4T)

VlnPlot(data, features = c("CD4T"), group.by = "seurat_clusters", pt.size=0, cols = soft_three_colors) + ylim(0,max.u)
ggsave("plots/cd4_violin_raw.pdf", height=2.5, width=3)

VlnPlot(data, features = c("CD4T_kNN"), group.by = "seurat_clusters", pt.size=0, cols = soft_three_colors) + ylim(0,max.u)
ggsave("plots/cd4_violin_knn.pdf", height=2.5, width=3)
```

Average by sample
```{r}
#Aggregate by sample and cluster
data$Sample_cluster <- paste(data$Sample, data$seurat_clusters, sep = "_c")

meta <- data@meta.data[,names(sigs),drop=F]
avgs <- lapply(unique(colnames(meta)), function(m) {
    tapply(meta[,m], data$Sample_cluster, mean)
})
names(avgs) <- colnames(meta)
avgs <- as.data.frame(avgs)

avgs$Sample_sub <- rownames(avgs)
melt.sig <- reshape2::melt(avgs)
colnames(melt.sig) <- c("Sample_cluster","Signature","value")

#Aggregate meta-features
library(dplyr)
library(tidyr)
melt.sig <- melt.sig %>% separate(Sample_cluster, c('Sample', 'Cluster'),
                                sep="_", remove=F)
```

Plot pseudo-bulks
```{r fig.height=5, fig.width=8}
#library(ggpubr)
library(patchwork)
which.sigs <- as.character(unique(melt.sig$Signature))

pll <- lapply(which.sigs, function(g) {
  max.u <- max(data@meta.data[,g])
  df.use <- melt.sig[melt.sig$Signature == g,]
  
  ggplot(df.use, aes(x=Cluster, y=value, color=Cluster, fill=NULL)) + coord_cartesian(ylim = c(0, max.u)) +
    geom_boxplot(outlier.shape = NA) + 
    geom_dotplot(binaxis='y', stackdir='center', dotsize=0.8,
                 position=position_dodge(0.8)) +
    theme_classic() + ggtitle(g) + scale_color_manual(values=soft_three_colors) 
#    stat_pwc(aes(group = Cluster), label = "p.signif", label.size=4, method = "wilcox_test", hide.ns = F)
})
names(pll) <- which.sigs

wrap_plots(pll)

```

```{r fig.height=3, fig.width=2.5}
pll[["melanocyte"]] + NoLegend() + theme(aspect.ratio = 1)
pll[["cytotox"]] + NoLegend() + theme(aspect.ratio = 1)
```

As jitterplots
```{r}
pll <- lapply(which.sigs, function(g) {
  max.u <- max(data@meta.data[,g])
  df.use <- melt.sig[melt.sig$Signature == g,]
  
  ggplot(df.use, aes(x=Cluster, y=value, color=Cluster, fill=NULL)) + coord_cartesian(ylim = c(0, max.u)) +
    geom_jitter(width = 0.3) +
    theme_classic() + ggtitle(g) + scale_color_manual(values=soft_three_colors) 
})
names(pll) <- which.sigs

for (i in 1:(length(pll)-1)) {
  pll[[i]] <- pll[[i]]
}
wrap_plots(pll)
```

```{r fig.height=3.5, fig.width=2}
pll[["melanocyte"]] + NoLegend() + theme(aspect.ratio = 1)

ggsave("plots/melanocyte_jitter.pdf", height=2, width=2.5)

pll[["cytotox"]] + NoLegend() + theme(aspect.ratio = 1)
```



# Simulate a condition (split patients into two groups)

```{r}
data$Condition = "Cond1"
data$Condition[data$Sample %in% c("LY31120","LY261021","LY021121")] <- "Cond2"
```

```{r}
cond_colors <- c("#E9C46A","#B22222")   
```

```{r fig.height=2.5, fig.width=3}
VlnPlot(data, features = c("cytotox_kNN"), group.by = "seurat_clusters", split.by = "Condition", pt.size=0, cols = cond_colors)
VlnPlot(data, features = c("melanocyte_kNN"), group.by = "seurat_clusters", split.by = "Condition", pt.size=0, cols = cond_colors)

ggsave("plots/melanocyte_violin_cond.pdf", height=2.5, width=3)
```


Average by sample
```{r}
#Aggregate by sample and cluster
data$Sample_cluster_cond <- paste(data$Sample, data$seurat_clusters, data$Condition, sep = "_")

meta <- data@meta.data[,names(sigs),drop=F]
avgs <- lapply(unique(colnames(meta)), function(m) {
    tapply(meta[,m], data$Sample_cluster_cond, mean)
})
names(avgs) <- colnames(meta)
avgs <- as.data.frame(avgs)

avgs$Sample_sub <- rownames(avgs)
melt.sig <- reshape2::melt(avgs)
colnames(melt.sig) <- c("Sample_cluster","Signature","value")

#Aggregate meta-features
library(dplyr)
library(tidyr)
melt.sig <- melt.sig %>% separate(Sample_cluster, c('Sample', 'Cluster', 'Condition'),
                                sep="_", remove=F)
```

Plot pseudo-bulks
```{r fig.height=5, fig.width=8}
#library(ggpubr)
library(patchwork)
which.sigs <- as.character(unique(melt.sig$Signature))

pll <- lapply(which.sigs, function(g) {
  max.u <- max(data@meta.data[,g])
  df.use <- melt.sig[melt.sig$Signature == g,]
  
  ggplot(df.use, aes(x=Cluster, y=value, color=Condition, fill=NULL)) + coord_cartesian(ylim = c(0, max.u)) +
    geom_boxplot(outlier.shape = NA) + 
    geom_dotplot(binaxis='y', stackdir='center', dotsize=0.8,
                 position=position_dodge(0.8)) +
    theme_classic() + ggtitle(g) + scale_color_manual(values=cond_colors) 
#    stat_pwc(aes(group = Cluster), label = "p.signif", label.size=4, method = "wilcox_test", hide.ns = F)
})
names(pll) <- which.sigs

for (i in 1:(length(pll)-1)) {
  pll[[i]] <- pll[[i]]
}
wrap_plots(pll)

```

```{r fig.height=3, fig.width=2.5}
pll[["melanocyte"]]  + theme(aspect.ratio = 2)
pll[["cytotox"]]  + theme(aspect.ratio = 2)
```

As jitterplots
```{r}
pll <- lapply(which.sigs, function(g) {
  max.u <- max(data@meta.data[,g])
  df.use <- melt.sig[melt.sig$Signature == g,]
  
  ggplot(df.use, aes(x=Cluster, y=value, color=Condition)) + coord_cartesian(ylim = c(0, max.u)) +
    geom_jitter(position = position_jitterdodge(jitter.width = 0.3, dodge.width = 0.8)) +
    theme_classic() + ggtitle(g) + scale_color_manual(values=cond_colors) 
})
names(pll) <- which.sigs

for (i in 1:(length(pll)-1)) {
  pll[[i]] <- pll[[i]]
}
wrap_plots(pll)
```

```{r fig.height=3, fig.width=4}
pll[["melanocyte"]]  + theme(aspect.ratio = 1.2)

ggsave("plots/melanocyte_jitter_cond.pdf", height=2, width=4.5)

pll[["cytotox"]]  + theme(aspect.ratio = 1.2)
```


Using a mixed-effect model
```{r}
library(lme4)
library(lmerTest)

df <- data@meta.data

model <- lmer(melanocyte ~ seurat_clusters * Condition + (1 | Sample), data = df)
summary(model)
```

```{r}
model <- lmer(melanocyte ~ seurat_clusters + (1 | Sample), data = df)
summary(model)
```

```{r}
library(emmeans)
library(ggplot2)

emm <- emmeans(model, ~ seurat_clusters)
emm_df <- as.data.frame(emm)

ggplot(emm_df, aes(x = seurat_clusters, y = emmean)) +
  geom_col(fill = "gray80", width = 0.6) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0.1) +
  labs(y = "Adjusted mean melanocyte UCell score", x = "Cluster") +
  theme_classic(base_size = 12)

df_summary <- df %>%
  group_by(Sample, seurat_clusters) %>%
  summarise(mean_score = mean(melanocyte), .groups = "drop")

```
